Higher-Order unification for free!

### 2024-04-15 paper/main.tex
sec 5.1 spellcheck

\section{Handling of \maybeeta}\label{sec:eta}
A term 
$\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
$x$ does not occur free in $t$, and conversely $t$ is the $\eta$-contraction of
$\lambda x. t \appsep x$. The equational theory of \Fo{} identifies these terms,
but the current compilation scheme does not,
as shown by the following example:
%
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}

@@

In order to guarantee \cref{prop:simulation} we detect
lambda abstractions that can disappear by $\eta$-contraction
(\cref{sec:etadetection}) and we modify the compiler so that it
generates fresh unification variables
in their place and moves the problematic term 
from \hoUnifPb to \linkStore (\cref{sec:etacomp}). The compilation
of the problem \foUnifPb above is refined to: 
%
\printAlll
  {{{\lambda x.X \appsep x,f}}}
  {{{A,f}}}
  {{{X,B,1}}}
  {{{\eta,,A,\lambda x.B_x}}}

\noindent
As per \cref{inv:linklhs} the \linketa left hand side is a variable
while the right hand side is a term in \maybeeta that has the following property:

@@

When compiling a term $t$ we need to determine if any
subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
The detection of lambda abstractions that can ``disappear''
is not as trivial as it may seems, here a few examples:
%
\begin{center}
  \begin{tabular}{lll}
    %Term & Status & Evidence \\\hline
    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x,~ B \mapsto \lambda y.\lambda x.y ~\}$
  \end{tabular}
\end{center}
\vspace{4pt}

% In the examples above, the first term is a \maybeeta since $A_x$ can
% reduce to $x$ by setting $A_x = \lambda x.x$, 
% the second one is not a \maybeeta since it exists no substitution
% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
\noindent
The first two examples are easy, and show how a unification variable can expose
or erase a variable in their scope and turn the resulting term in an $\eta$-expansion or not.\\
The third example shows that when a variable occurs outside the scope of a unification
variable it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
The last example shows the recursive nature of the check we need to implement.
The term starts with a spine of two lambdas hence the whole term
is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
is in \maybeeta itself. If it is, it could $\eta$-contract to
$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
$\eta$-expansion.\\
We can now define more formally how \maybeeta terms are detected together with
its auxiliary functions:

@@

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w_i \in W$, such that $w_i$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case the term $t$ is under the spine of binders
$x_1\ldots x_n$, $t$ can either be $x$ applied to terms that can
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to that application (case 3).
\end{proof}

@@

In other words $x$ \occursrigid in $t$ if it occurs in $t$
outside of the scope of a unification variable $X$, otherwise an instantiation
of $X$ can make $x$ disappears from $t$.
Moreover, note that $\eta$-contracting $t$ cannot make $x$ disappear, since
$x$ is not a locally bound variable inside $t$.

@@

As we will see in the rest of this section this is not a problem
since it does not break
\cref{prop:simulation} nor \cref{prop:fidelity}.
% A term in \maybeeta{} is compiled to a
% unification variable and a link (see \cref{sec:etacomp}):
% the link makes progress (see \cref{sec:etaprogress})
% in the same step in which the
% variable is instantiated, and that compensates
% for this coarse analysis.

>>>>

\section{Handling of \maybeeta}\label{sec:eta}
A term 
$\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
$x$ does not occur free in $t$, and conversely, $t$ is the $\eta$-contraction of
$\lambda x. t \appsep x$. The equational theory of \Fo{} identifies these terms,
but the current compilation scheme does not,
as shown by the following example:
%
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}

@@

In order to guarantee \cref{prop:simulation}, we detect
lambda abstractions that can disappear by $\eta$-contraction
(\cref{sec:etadetection}), and we modify the compiler so that it
generates fresh unification variables
in their place and moves the problematic term 
from \hoUnifPb to \linkStore (\cref{sec:etacomp}). The compilation
of the problem \foUnifPb above is refined to: 
%
\printAlll
  {{{\lambda x.X \appsep x,f}}}
  {{{A,f}}}
  {{{X,B,1}}}
  {{{\eta,,A,\lambda x.B_x}}}

\noindent
As per \cref{inv:linklhs} the \linketa left-hand side is a variable
while the right-hand side is a term in \maybeeta that has the following property:

@@

When compiling a term $t$, we need to determine if any
subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
The detection of lambda abstractions that can ``disappear''
is not as trivial as it may seems. Here a few examples:
%
\begin{center}
  \begin{tabular}{lll}
    %Term & Status & Evidence \\\hline
    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x,~ B \mapsto \lambda y.\lambda x.y ~\}$
  \end{tabular}
\end{center}
\vspace{4pt}

% In the examples above, the first term is a \maybeeta since $A_x$ can
% reduce to $x$ by setting $A_x = \lambda x.x$, 
% the second one is not a \maybeeta since it exists no substitution
% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
\noindent
The first two examples are easy, and show how a unification variable can expose
or erase a variable in its scope, turning the resulting term into an $\eta$-expansion or not.\\
The third example shows that when a variable occurs outside the scope of a unification
variable, it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
The last example shows the recursive nature of the check we need to implement.
The term starts with a spine of two lambdas, hence the whole term
is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
is in \maybeeta itself. If it is, it could $\eta$-contract to
$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
$\eta$-expansion.\\
We can now define more formally how \maybeeta terms are detected together with
its auxiliary functions:

@@

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w_i \in W$, such that $w_i$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case, the term $t$ is under the spine of binders
$x_1\ldots x_n$, $t$ can either be $x$ applied to terms that can
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to that application (case 3).
\end{proof}

@@

In other words, $x$ \occursrigid in $t$ if it occurs in $t$
outside of the scope of a unification variable $X$; otherwise, an instantiation
of $X$ can make $x$ disappears from $t$.
Moreover, note that $\eta$-contracting $t$ cannot make $x$ disappear, since
$x$ is not a locally bound variable inside $t$.

@@

As we will see in the rest of this section, this is not a problem
since it does not break
\cref{prop:simulation} nor \cref{prop:fidelity}.
% A term in \maybeeta{} is compiled to a
% unification variable and a link (see \cref{sec:etacomp}):
% the link makes progress (see \cref{sec:etaprogress})
% in the same step in which the
% variable is instantiated, and that compensates
% for this coarse analysis.
